The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015

### 2017-09-04 paper/paper.Rnw
Produced first two tables from tracking modeling - everything works after restart

<<eval = F>>=

@@

# 2015
# SC013Q01TA
# 1 Public
# 2 Private
# 7 N/A
# 8 Invalid
# 9 Missing
library(readxl)

@@

library(inequalityintsvy)

central_examination <-
  tribble(
  ~ country, ~central_examination,
  "Australia", 1, 
  "Austria", 0, 
  "Belgium", 0,
  "Bulgaria", 1,
  "Canada", rbinom(1, 1, prob = 0.51),
  "Czech Republic", 1,
  "Denmark", 1,
  "Finland", 1,
  "France", 1,
  "Germany", 0,
  "United Kingdom", 1,
  "Greece", 0,
  "Hong Kong", 1,
  "Hungary", 1,
  "Iceland", 1,
  "Ireland", 1,
  "Israel", 1,
  "Italy", 1,
  "Japan", 1,
  "Korea", 1,
  "Latvia", 1,
  "Liechtenstein", 1,
  "Luxembourg", 1,
  "Netherlands", 1,
  "New Zealand", 1,
  "Norway", 1,
  "Poland", 1,
  "Portugal", 0,
  "Russia", 1,
  "Slovakia", 1,
  "Slovenia", 1,
  "Spain", 0,
  "Sweden", 0,
  "Switzerland", 0,
  "Turkey", 1,
  "United States", 1
  )

# The tracking variables explain very nicely the difference in test scores
# for all countries pooled, up to 50% R squared.

@@

gaps <- c("90th/10th SES gap", "80th/20th SES gap", "70th/30th SES gap")

all_gaps_models <-
  map(gaps, function(gap) {
    ready_data_age <-
      dif_data(complete_gaps, tracking_data, central_examination, which_gap = gap) %>%
      mutate(age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor)

@@

model_data_merged %>%
  ggplot(aes(terms, estimate, colour = model)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_colour_discrete(name = NULL) +
  scale_y_continuous(limits = c(-0.5, 2)) +
  ggtitle("Coef from linear model of inequality on degree of tracking and vocational tracking") +
  coord_flip()

library(brms)
# It under predicts the y value rather than over predict
# It has problems predicting the central tendency of the model
# This linear model explains the pooled 90/10 differences

@@

pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
pp_check(mod_tracking, type = "scatter_avg", nsamples = 100)
pp_check(mod_tracking, type = "boxplot", nsamples = 20)
pp_check(mod_tracking, type = "error_hist", nsamples = 6)
pp_check(mod_tracking, type = "ecdf_overlay")

pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "zvoc")
pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "ztrack")

# I don't know how to estimate VIF's with bayesian reasoning
model_formula <- difference ~ num_tracks + age_selection + zvoc + I(length > 0)
mod_tracking_ext <-
  brms::brm(model_formula, family = gaussian(),
  data = ready_data_age,
  warmup = 1000, iter = 2000, chains = 5
  )

bayes_R2(mod_tracking_ext)

# This is the closest I could get to getting vif's:
lm(model_formula, data = ready_data_age) %>%
  vif()

new_data <-
  ready_data_age %>%
  with(crossing(country = countries,
                num_tracks = unique(num_tracks),
                age_selection = unique(age_selection),
                zvoc = mean(zvoc, na.rm = TRUE),
                length = typical(length > 0)))

new_data %>%
  left_join(sum_ready_data) %>%
  ggplot(aes(num_tracks %>% as.factor, avg_diff, fill = num_tracks)) +
  geom_errorbar(aes(ymin = diff_low, ymax = diff_high), width = 0.2, colour = "red") +
  geom_point(colour = "red") +
  scale_y_continuous(limits = c(0, 4)) +
  geom_col(aes(y = pred, alpha = 0.4)) +
  geom_errorbar(aes(ymin = pred_low, ymax = pred_high), width = 0.2) +
  facet_wrap(country ~ age_selection) +
  scale_fill_wsj()
@

>>>>

<<first_model_table>>=

@@

# 2015
# SC013Q01TA
# 1 Public
# 2 Private
# 7 N/A
# 8 Invalid
# 9 Missing
library(readxl)
library(inequalityintsvy)
library(brms)

@@

# central_examination <-
#   tribble(
#   ~ country, ~central_examination,
#   "Australia", 1, 
#   "Austria", 0, 
#   "Belgium", 0,
#   "Bulgaria", 1,
#   "Canada", rbinom(1, 1, prob = 0.51),
#   "Czech Republic", 1,
#   "Denmark", 1,
#   "Finland", 1,
#   "France", 1,
#   "Germany", 0,
#   "United Kingdom", 1,
#   "Greece", 0,
#   "Hong Kong", 1,
#   "Hungary", 1,
#   "Iceland", 1,
#   "Ireland", 1,
#   "Israel", 1,
#   "Italy", 1,
#   "Japan", 1,
#   "Korea", 1,
#   "Latvia", 1,
#   "Liechtenstein", 1,
#   "Luxembourg", 1,
#   "Netherlands", 1,
#   "New Zealand", 1,
#   "Norway", 1,
#   "Poland", 1,
#   "Portugal", 0,
#   "Russia", 1,
#   "Slovakia", 1,
#   "Slovenia", 1,
#   "Spain", 0,
#   "Sweden", 0,
#   "Switzerland", 0,
#   "Turkey", 1,
#   "United States", 1
#   )

@@

ivs <- c("num_tracks", "age_selection", "I(length > 0)", "zvoc")

formula_seq <-
  lapply(1:length(ivs), function(x) {
    paste0("difference ~ ", paste0(ivs[1:x], collapse = " + ") %>% paste0(" + (1|year)"))
  }) %>%
  lapply(as.formula)

all_track_models <- 
  map(formula_seq, ~ {

  mod_tracking <-
  brms::brm(
    .x, family = gaussian(),
    data = ready_data_age,
    warmup = 1000, iter = 2000, chains = 5,
    control = list(adapt_delta = 0.90)
  )

  mod_tracking

})

tidy_dfs <-
  map(all_track_models, ~ {

  broom::tidy(.x) %>%
    .[grepl("^b_", .$term), ] %>%
    map_if(is_double, round, 2) %>%
    as_tibble %>%
    transmute(term,
              estimate = paste0(estimate, " (", lower, "/", upper, ")"))
  })

tidy_dfs_two <-
  reduce(tidy_dfs, full_join, by = "term") %>%
  setNames(c(" ", paste("Model", 1:4))) %>%
  mutate(" " = c(
    "Intercept",
    "Num of tracks = 1",
    "Age at track >= 15",
    "% curriculum tracked",
    "Vocational Index"
  )) %>%
  .[c(2:5, 1), ]

r_square <-
  paste0("\\hline ", paste0("R square: & ",
         paste0(map(all_track_models, ~ bayes_R2(.x)[1] %>% round(2) * 100) %>% paste0("\\%"),
                       collapse = " & ")), " \\\\")
n_obs <-
  paste0(paste0(" Sample size: & ",
         paste0(map(all_track_models, nobs), collapse = " & ")), " \\\\ \\hline")

command_to_row <- paste0(r_square, n_obs)

addtorow_two <- list(pos = list(pos = 5), command = command_to_row)

xtable(tidy_dfs_two) %>% print.xtable(
  include.rownames = FALSE,
  add.to.row = addtorow_two
  )
@

<<>>=
# This is the closest I could get to getting vif's fro the last model:

lm(difference ~ num_tracks + age_selection + I(length > 0) + zvoc,
   data = ready_data_age) %>%
  vif()

mod_tracking_ext <-
  brms::brm(formula_seq[[4]],
            family = gaussian(),
            data = ready_data_age,
            warmup = 1000, iter = 2000, chains = 5,
            control = list(adapt_delta = 0.90)
  )

new_data <-
  ready_data_age %>%
  with(crossing(country = countries,
                num_tracks = unique(num_tracks),
                age_selection = unique(age_selection),
                zvoc = mean(zvoc, na.rm = TRUE),
                length = typical(length > 0),
                year = "2009"))

new_data %>%
  left_join(sum_ready_data) %>%
  mutate(age_selection = ifelse(age_selection == 1, ">= 15", "< 15")) %>%
  ggplot(aes(num_tracks, avg_diff, fill = num_tracks)) +
  geom_errorbar(aes(ymin = diff_low, ymax = diff_high), width = 0.2, colour = "red") +
  geom_point(colour = "red") +
  geom_col(aes(y = pred, alpha = 0.4)) +
  geom_errorbar(aes(ymin = pred_low, ymax = pred_high), width = 0.2) +
  facet_wrap(country ~ age_selection) +
  scale_y_continuous(name = "90/10 SD gap", limits = c(0, 4)) +
  scale_alpha(guide = F) +
  scale_x_discrete(name = "Number of tracks", labels = c("> 1 track", "1 track")) +
  scale_fill_manual(values = c("orangered2", "blue3"), guide = F) +
  ggtitle("Predicted 90/10 gap for different combinations of number of tracks and age of selection")
@

<<>>=
pred_avg_diff<-
  new_data %>%
  left_join(sum_ready_data) %>%
  filter(!is.na(avg_diff) | num_tracks == 1 & age_selection == 1) %>%
  select(country, num_tracks, age_selection, pred, avg_diff)

track_should_change <- pred_avg_diff$country[pred_avg_diff$country %>% duplicated]

avg_diff_data <-
  pred_avg_diff %>%
  filter(country %in% track_should_change) %>%
  transmute(country,
            avg_diff = ifelse(avg_diff %>% is.na, pred, avg_diff),
            combn = paste0(num_tracks, "_", age_selection)) %>%
  filter(combn != "1_1")

final_avg_diff_table <-
  avg_diff_data %>%
  left_join(pred_avg_diff %>%
              filter(country %in% track_should_change & (num_tracks == 1 & age_selection == 1)) %>%
              select(country, pred), by = "country") %>%
  select(country, combn, avg_diff, pred) %>%
  separate(combn, c("num_tracks", "age_selection"), sep = "_") %>%
  transmute(
    Country = country,
    `Number of tracks` = ifelse(num_tracks == 0, "More than one track", "One track"),
    `Age of selection` = ifelse(age_selection == 1, "15 or more", "Less than 15"),
    `Actual inequality` = avg_diff %>% round(3),
    `Predicted inequality` = pred %>% round(3),
    `% reduction` = (((`Actual inequality` - `Predicted inequality`)/ `Actual inequality`) * 100) %>%
      ceiling %>% paste0("%")
    )

xtable(final_avg_diff_table) %>% print.xtable(
  include.rownames = FALSE
  )

gaps <- c("90th/10th SES gap", "80th/20th SES gap", "70th/30th SES gap")
all_gaps_models <-
  map(gaps, function(gap) {
    ready_data_age <-
      dif_data(complete_gaps, tracking_data, central_examination, which_gap = gap) %>%
      mutate(age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor)

    mod_tracking <-
      brms::brm(
        difference ~
          ztrack +
          zvoc +
          ztrack:zvoc +
          (1|year), family = gaussian(),
      data = ready_data_age,
      warmup = 1000, iter = 2000, chains = 5,
      control = list(adapt_delta = 0.95)
      )

@@

model_data_merged %>%
  ggplot(aes(terms, estimate, colour = model)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_colour_discrete(name = NULL) +
  scale_y_continuous(limits = c(-0.5, 2)) +
  scale_x_discrete(labels = c("Intercept",
                              "Tracking Index",
                              "Vocational * Tracking Index",
                              "Vocational Index")) +
  labs(y = "Estimate from linear model", x = "Predictors") +
  ggtitle("Estimation of tracking/vocational models for different gaps") +
  coord_flip()

# It under predicts the y value rather than over predict
# It has problems predicting the central tendency of the model
# This linear model explains the pooled 90/10 differences

@@

plot(plot_interaction, plot = FALSE)[[1]] +
  labs(x = "Tracking Index (SD)", y = "Difference in test scores (SD)") +
  scale_colour_discrete(
    name = "Vocational index \n quantiles",
    labels = seq(100, 0, -25) %>% paste0("%")
    ) +
  scale_fill_discrete(guide = FALSE) +
  ggtitle("Interaction between Tracking and Vocational index")

# pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
# pp_check(mod_tracking, type = "scatter_avg", nsamples = 100)
# pp_check(mod_tracking, type = "boxplot", nsamples = 20)
# pp_check(mod_tracking, type = "error_hist", nsamples = 6)
# pp_check(mod_tracking, type = "ecdf_overlay")
# 
# pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "zvoc")
# pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "ztrack")
